---
title: 'MoveR: an R package to ease animal movement analyses'
author: "Quentin PETITJEAN"
date: "05/07/2022"
output: 
  html_document:
    toc: TRUE
    toc_depth: 3
    toc_float: TRUE
    number_sections: TRUE
    use_fontawesome: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries 
```{r libraries, eval = FALSE}
if(require(remotes)) {
  install.packages("remotes")
}
library("remotes")
if(require(MovR)) {
  remotes::install_github("qpetitjean/MovR")
}
library("MovR")
```

# Load the exemple dataset (Trex tracker)
```{r LoadDataset}
# path to the video file (replace by the github repository - once open source)
videoPath = "E:/Videos_comportement/Bidime_Trex_npzfiles/2021-03-01-ISA17150-chaud"
# path to the arena border file (generated using imageJ) (replace by the github repository - once open source)
ArenaFile = "D:/Postdoc_INRAE_SAM/Data_Bidime/Scale_IndCount_Bidime/Img_video/2021-03-01-ISA17150-chaud/DSC_8113.MOV_1.txt"

# the scaling of the video - 1 cm represent 413.4 pixels
scaling = 1 / 413.4

# the frame rate of the video expressed in frame per second
frameRate = 25

# use the readTrex function from MovR package to load the tracking results
Data_Trex <- MovR::readTrex(trexPath = videoPath,
                            imgHeight = 1080,
                            mirrorY = TRUE)
str(Data_Trex)
```
Here, the results of the tracking software (i.e., Trex) are loaded using the "read" function (see readTrex(), readCtrax(), readIdtracker(), readTrackR()). The function load the bunch of .npz file, retrieve the useful informations and lay them in a list (for informations about the informations returned see ??readTrex()). Using the optional argument "rawDat", the user can have access to all the data stored into the .npz files (see ??readTrex()). Also, since Trex use the upper left corner of the image as origin, the x and y coordinates are mirrored to start at the lower left corner using the "mirrorY" argument.

```{r ConvertDataset}
# remove 40 last minutes of the video to accelerate computation
Data_Trex_Sample <-
  as.data.frame(Data_Trex[1:7])[which(Data_Trex[["frame"]] <= 60 * 60 * 25), ]

# convert the data to a list of fragment
trackDat <- MovR::convert2frags(Data_Trex_Sample,
                                by = "identity")
```
To ease further computation and graphical representation the data are then converted to a list of fragments (based on particle' identity), corresponding to the trajectory of all the particles detected over the tracking, using the "convert2frags" function. 

# clean the dataset 

Once loaded, the data need to be cleaned to remove for instance, infinite values due to moments where particles were undetected, or the detection of spurious elements during the tracking. For this purpose, 2 main functions allow the user to define custom filters and apply them to the dataset.

## remove infinite values 

Infinite values are added when the particles are temporarily undetected, the following code help to remove infinite value and split the fragment carrying them accordingly (when a particle is temporarily undetected we assume that the identity of the fragment is spurious, thus the function create a new fragment).

```{r InfClean}
# define the filter to detected infinite values on "x.pos"
filter.Inf <-
  filterFunc(
    trackDat,
    toFilter = "x.pos",
    customFunc = function(x)
      is.infinite(x)
  )

# filter infinite values according to the previously specified filter
# NB: Since the dataset correspond to the tracking of particles over a period of 60 minutes, we remove the fragments that are shorter than 100 frames (about 1% of the total video duration) using "mindur" argument.
trackDat.Infilt <-
  filterFrags(trackDat,
              filter = filter.Inf,
              splitCond = TRUE,
              minDur = 100)

# the output of the function correspond to a list, with information about the filtration process (see ??filterFrags())
str(trackDat.Infilt[[1]])

# while the second element of the list contains the list of filtered fragment which can be used for further computation

```

## Filter based on particles size 

Spurious particles can be detected during the tracking procedure. One cleaning step can consist on removing all the moment when a particles' size is lower or higher than a given threshold.
Here we keep only the moments where a particle' size is included within the 95% confidence interval based on the distribution of the particle' size over the dataset.

```{r SizeClean}
# convert the previously filtered data into a list to ease representation and computation of 95% IC of the particles' size distribution
trackDat.InfiltList <- convert2list(trackDat.Infilt[[2]])

# compute IC
## transform particles' size in log10 to approximate the normal distribution
indLength <- log10(trackDat.InfiltList$maj.ax)
if (length(which(is.infinite(indLength)) > 0)) {
  indLength <- indLength[-c(which(is.infinite(indLength)))]
}
if (length(which(is.na(indLength)) > 0)) {
  indLength <- indLength[-c(which(is.na(indLength)))]
}
IC <- quantile(indLength, c(0.025, 0.975))

# plot the resulting distribution and the 95% IC
hist(log10(trackDat.InfiltList$maj.ax),
     breaks = 100,
     main = "Indiv length (log10) and 95% IC")
abline(v = c(IC[1], IC[2]))

# the result seem satisfactory
# hence, create the filter based on the computed IC values
filter.length <-
  filterFunc(
    trackDat.Infilt[[2]],
    toFilter = "maj.ax",
    customFunc = function(x)
      x < 10 ^ IC[1] | x > 10 ^ IC[2]
  )

# apply the filter on the data 
trackDat.lenfilt <-
  filterFrags(trackDat.Infilt[[2]],
              filter.length,
              splitCond = TRUE,
              minDur = 100)

# display the information about the filtration process (see ??filterFrags())
str(trackDat.lenfilt[[1]])
```

## Filter based on particles speed 

While the data can be filtered based data that already exist into the tracking result, it is also possible to compute new metrics and filter the dataset based on it.
For instance, particles' speed can be a good indicator of spurious detection or particle identification. Indeed, when the speed of a particle is too high (and have hence no biological meaning) it can be due to change in particle identity or more generally tracking artifact. 

```{r SpeedClean}
# define filters based on the 999th percentile of the individuals speed (not listed in the raw tracking data, need some computation)

# retrieve the previously filtered data and used them to compute particles' speed
trackDat2 <- trackDat.lenfilt[[2]]

# to compute particles' speed we use another function from the MoveR package, the "analyseFrags" function which is used to performed any computation (either manually specified by the user or using already existing modulus such as speed()) along each fragment present in the dataset. 
# here we use the "speed()" modulus to compute the speed of every particles' trajectories
trackDat2 <-
  analyseFrags(trackDat2,
               customFunc = list(
                 speed = function(x)
                   MovR::speed(
                     x,
                     scale = scaling,
                     unit = "cm",
                     TimeCol = "frame"
                   )
               ))

# compute the 999th quantile 
# convert the previously filtered data into a list to ease representation and computation of 999th quantile of the particles' speed distribution
trackDat.speedfiltList <- convert2list(trackDat2)

## transform particles' speed in log10 to approximate the normal distribution
indSpeed <- log10(trackDat.speedfiltList$speed)
if (length(which(is.infinite(indSpeed)) > 0)) {
  indSpeed <- indSpeed[-c(which(is.infinite(indSpeed)))]
}
if (length(which(is.na(indSpeed)) > 0)) {
  indSpeed <- indSpeed[-c(which(is.na(indSpeed)))]
}
quant999th <- quantile(indSpeed, c(0.999))

# plot the resulting distribution and the 999th quantile 
hist(indSpeed, breaks = 100, main = "Indiv speed (log10) and 999th quantile")
abline(v = quant999th)

# the result seem satisfactory
# hence, create the filter base on the computed 999th quantile 
filter.speed <-
  filterFunc(
    trackDat2,
    toFilter = "speed",
    customFunc = function(x)
      x < 0 | x > 10 ^ quant999th
  )

# apply the filter on the data 
trackDat.speedfilt <-
  filterFrags(trackDat2,
              filter.speed,
              splitCond = TRUE,
              minDur = 100)

# display the information about the filtration process (see ??filterFrags())
str(trackDat.speedfilt[[1]])

```

## Filter based on particle detection outside the arena

Sometimes particles can be detected outside the arena, either because an individual as escaped from an arena or because the background has changed over the video recording (tracking artifact). To solve this issue, it is possible to remove the particles that have been detected outside the arena.

```{r edgeClean}
# define filter based on the presence of individual outside the arena

# load gradient to the edge of the arena (a distance matrix generated using color tresholding in imageJ) 
arenaGrad <- read.delim(ArenaFile)

# represent the gradient of distance to the arena edge
contour(as.matrix(arenaGrad))

# The edge correspond to the lower value of the gradient (i.e., 1), 
# we can hence extract the arena edge from the distance matrix as follow
edge <- data.frame(which(arenaGrad == 1, arr.ind = T))
names(edge)[c(1, 2)] <- c("y.pos", "x.pos")
plot(NA, xlim = c(0, 1920), ylim = c(0, 1080))
points(x = edge[, "x.pos"], y = edge[, "y.pos"], cex = 0.01)

# As well as the center of the arena
center = c(mean(edge[, "x.pos"]), mean(edge[, "y.pos"]))
points(
  x = center[1],
  y = center[2],
  col = "black",
  pch = 3,
  cex = 0.5
)

# As the dataset is relatively large, we will represent a subset of the data
## retrieve the maximum value of time within the dataset
maxFrame <-
  max(convert2list(trackDat.speedfilt[[2]])$frame, na.rm = T)

## specify a custom timeline to draw  1000 frame every 20000 frames
tempTimeline <-  lapply(sort(c(
  seq(0, maxFrame / 1000)[seq(1, length(seq(maxFrame / 1000)), by = 20)],
  seq(0, maxFrame / 1000)[seq(2, length(seq(maxFrame / 1000)), by = 20)]
)), function(x)
  c(x * 1000, (x + 1) * 1000))
Timeline <-
  lapply(seq(1, length(tempTimeline), by = 2), function(x)
    tempTimeline[[x]])

## plot fragments with the arena to see whether some part are detected outside the arena
drawFrags(
  trackDat.speedfilt[[2]],
  imgRes = c(1920, 1080),
  timeWin = Timeline,
  timeCol = "frame",
  add2It = list(
    points(x = edge[, "x.pos"], y = edge[, "y.pos"], cex = 0.01),
    points(
      x = center[1],
      y = center[2],
      col = "black",
      pch = 3,
      cex = 0.5
    )
  )
)

# compute the distance to the edge using "analyseFrag" and the dist2edge modulus
trackDat3 <- trackDat.speedfilt[[2]]
trackDat3 <-
  analyseFrags(trackDat3,
               customFunc = list(
                 dist2Edge = function(x)
                   MovR::dist2Edge(x, edge,
                                   customFunc = "CircularArena")
               ))

# represent the proportion of particles detected outside the arena
# here we consider that an individual is truly detected
# outside when more than half of the mean body length is out of the arena
## extract the vector of distance to the edge
indOut <- convert2list(trackDat3)$dist2Edge

## compute the half of the mean body length of the particles
meanBodyL <-
  mean(convert2list(trackDat3)$maj.ax, na.rm = T) / 2

## plot it
hist(indOut, breaks = 100, main = "Distance to the edge, \npositive values correspond to Indivs that are outside the arena")
abline(v = meanBodyL)

# the result seem satisfactory 
# hence, create the filter accordingly
filter.out <-
  filterFunc(
    trackDat3,
    toFilter = "dist2Edge",
    customFunc = function(x)
      x > meanBodyL
  )

# apply the filter on the data 
trackDat.outfilt <-
  filterFrags(trackDat3,
              filter.out,
              splitCond = TRUE,
              minDur = 100)

# display the information about the filtration process (see ??filterFrags())
str(trackDat.outfilt[[1]])

# rename the cleaned dataset for further use
trackDat4 <- trackDat.outfilt[[2]]

```

## Cleaning summary

Once the cleaning seem done, the various filter summary can be retrieved and easily grouped to keep them for the sacks of reproducibility.

It is also possible to display a summary of the tracking information of both the initial dataset (i.e., trackDat) and the cleaned one (trackDat4) using the "trackStats" function to quickly compare them.

```{r CleaningSummary}

# create a summary of each filter results 
FilterSummary <- do.call("cbind",
                         list(
                           data.frame(Infilt = unlist(trackDat.Infilt[[1]])),
                           data.frame(lenfilt = unlist(trackDat.lenfilt[[1]])),
                           data.frame(speedfilt = unlist(trackDat.speedfilt[[1]])),
                           data.frame(outfilt = unlist(trackDat.outfilt[[1]]))
                         ))

# compute and display a detailed summary of the tracking information before and after cleaning (see ??trackStats())
Data_Trex_stats_before_filter <-
  trackStats(trackDat,
             frameR = frameRate,
             scale = scaling,
             unit = "cm")

Data_Trex_stats_after_filter <-
  trackStats(trackDat4,
             frameR = frameRate,
             scale = scaling,
             unit = "cm")

str(Data_Trex_stats_before_filter)
str(Data_Trex_stats_after_filter)
```



